---
title: "Mortality among women in Brazil"
subtitle: "A descriptive analysis of homicide, suicide and femicide, 2010–2024"
author: "Felipe Barletta"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
    highlight: tango
    theme: united
    css: style.css
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = NA,
  cache = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  dpi = 160,
  out.width = "100%"
)

options(
  scipen = 999,
  dplyr.summarise.inform = FALSE
)

setwd("~/Documents/DataSus/Mulheres")

library(microdatasus)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(forcats)
library(readr)
library(readxl)
library(ggplot2)
library(scales)
library(cowplot)

# Study period.
year_start <- 2010
year_end   <- 2024
years      <- year_start:year_end

# Project directories.
data_directory       <- "data"
population_directory <- file.path(data_directory, "population")
output_directory     <- "results_women"
figure_directory     <- file.path(output_directory, "figures")

dir.create(data_directory, showWarnings = FALSE, recursive = TRUE)
dir.create(population_directory, showWarnings = FALSE, recursive = TRUE)
dir.create(output_directory, showWarnings = FALSE, recursive = TRUE)
dir.create(figure_directory, showWarnings = FALSE, recursive = TRUE)

# Sober plotting palette.
col_navy  <- "#0B4F6C"
col_blue  <- "#0077B6"
col_teal  <- "#00A6A6"
col_red   <- "#E85D5D"
col_gold  <- "#D89B2B"
col_grey  <- "#6C757D"

# Common plotting theme.
theme_datasus <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(
        face = "bold",
        size = base_size + 3,
        color = col_navy,
        margin = margin(b = 6)
      ),
      plot.subtitle = element_text(
        size = base_size,
        color = "#45555F",
        margin = margin(b = 10)
      ),
      plot.caption = element_text(
        size = base_size - 3,
        color = "#5D6A72",
        hjust = 0,
        lineheight = 1.2,
        margin = margin(t = 10)
      ),
      axis.title = element_text(
        face = "bold",
        color = "#334E5C"
      ),
      axis.text = element_text(
        color = "#45555F"
      ),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(
        color = "#E5EDF2",
        linewidth = 0.4
      ),
      legend.position = "top",
      legend.title = element_blank(),
      plot.margin = margin(10, 12, 10, 10)
    )
}

# Contact footer used in publication and social-media figures.
# The PNG files are stored locally in icons/ so the project remains reproducible.
contact_footer <- function() {
  stopifnot(
    file.exists("icons/gmail.png"),
    file.exists("icons/whatsapp.png")
  )

  cowplot::ggdraw() +
    cowplot::draw_label(
      "Analysis and interpretation: Felipe Barletta",
      x = 0.01, y = 0.50,
      hjust = 0, vjust = 0.5, size = 9
    ) +
    cowplot::draw_image(
      "icons/gmail.png",
      x = 0.39, y = 0.34,
      width = 0.025, height = 0.32
    ) +
    cowplot::draw_label(
      "e-mail.felipe.barletta@gmail.com",
      x = 0.42, y = 0.50,
      hjust = 0, vjust = 0.5, size = 9
    ) +
    cowplot::draw_label(
      "|",
      x = 0.675, y = 0.50,
      hjust = 0.5, vjust = 0.5, size = 9
    ) +
    cowplot::draw_image(
      "icons/whatsapp.png",
      x = 0.70, y = 0.34,
      width = 0.025, height = 0.32
    ) +
    cowplot::draw_label(
      "+351 910 678 043",
      x = 0.73, y = 0.50,
      hjust = 0, vjust = 0.5, size = 9
    )
}

add_contact_footer <- function(plot_object) {
  cowplot::plot_grid(
    plot_object,
    contact_footer(),
    ncol = 1,
    rel_heights = c(1, 0.075),
    align = "v",
    axis = "lr"
  )
}
```

# Study overview

This project uses official mortality microdata from the Brazilian **Mortality Information System (SIM/DATASUS)** to describe mortality among women in Brazil from 2010 to 2024. SIM is based on information recorded on the Brazilian **Declaração de Óbito (Death Certificate)**, including demographic characteristics and the underlying cause of death coded according to ICD-10.

The analysis follows a deliberately progressive narrative: **all female deaths → population-based mortality → external causes → homicide and suicide → femicide**. Additional age and race/skin-color analyses are retained at the end for subsequent analyses and communication products.

Official sources:

- [SIM — Brazilian Ministry of Health](https://www.gov.br/saude/pt-br/composicao/svsa/sistemas-de-informacao/sim)
- [DATASUS](https://datasus.saude.gov.br/)
- [IBGE Population Projections, 2024 Revision](https://www.ibge.gov.br/estatisticas/sociais/populacao/9109-projecao-da-populacao.html)
- [Fórum Brasileiro de Segurança Pública — Anuário Brasileiro de Segurança Pública](https://forumseguranca.org.br/publicacoes/anuario-brasileiro-de-seguranca-publica/)

## Cohort definition

The analytical cohort contains deaths recorded as **female** in SIM between 2010 and 2024. It is not restricted to deaths occurring in hospitals and is not restricted by occupation, employment status, education, marital status or place of occurrence.

The principal SIM variables used are date of death (`DTOBITO`), underlying cause (`CAUSABAS`), sex (`SEXO`), age (`IDADE`), race/skin color (`RACACOR`) and municipality of residence (`CODMUNRES`).

# Reproducible data preparation

## Download SIM and identify women

Each year is downloaded and filtered separately. The cached RDS file is reused when available.

```{r download-data, results='hide'}
sim_variables <-
  c(
    "DTOBITO",
    "CAUSABAS",
    "SEXO",
    "IDADE",
    "RACACOR",
    "CODMUNRES"
  )

download_female_deaths <- function(year) {

  sim_raw <-
    fetch_datasus(
      year_start = year,
      year_end = year,
      uf = "all",
      information_system = "SIM-DO",
      vars = sim_variables,
      stop_on_error = TRUE,
      timeout = 600,
      quiet = FALSE
    )

  # In raw SIM-DO data, SEXO = 2 identifies female records.
  sim_raw <-
    sim_raw %>%
    filter(
      str_trim(as.character(SEXO)) == "2"
    )

  sim_processed <-
    process_sim(
      sim_raw,
      municipality_data = FALSE
    )

  women_year <-
    sim_processed %>%
    mutate(YEAR = year)

  rm(sim_raw, sim_processed)
  gc()

  women_year
}

women_rds <-
  file.path(
    data_directory,
    "female_deaths_SIM_2010_2024.rds"
  )

if (file.exists(women_rds)) {

  women_deaths <- readRDS(women_rds)

} else {

  women_deaths <-
    purrr::map_dfr(
      years,
      download_female_deaths
    )

  saveRDS(
    women_deaths,
    women_rds,
    compress = TRUE
  )
}
```

## Create analysis variables

```{r create-variables, results='hide'}
women_deaths <-
  women_deaths %>%
  mutate(
    death_date =
      as.Date(
        DTOBITO,
        format = "%Y-%m-%d"
      ),

    death_year =
      as.integer(
        format(
          death_date,
          "%Y"
        )
      ),

    age =
      suppressWarnings(
        as.numeric(IDADEanos)
      ),

    underlying_cause =
      str_to_upper(
        str_trim(
          as.character(CAUSABAS)
        )
      ),

    icd10_3 =
      str_sub(
        underlying_cause,
        1,
        3
      ),

    age_group =
      cut(
        age,
        breaks = c(
          -Inf, 19, 29, 39, 49,
          59, 69, 79, Inf
        ),
        labels = c(
          "<20", "20–29", "30–39", "40–49",
          "50–59", "60–69", "70–79", "80+"
        ),
        right = TRUE
      ),

    # Mutually exclusive external-cause categories.
    external_cause_category =
      case_when(
        str_detect(icd10_3, "^V") ~
          "Transport accidents",

        icd10_3 >= "W00" & icd10_3 <= "W19" ~
          "Falls",

        icd10_3 >= "W65" & icd10_3 <= "W74" ~
          "Drowning",

        icd10_3 >= "X40" & icd10_3 <= "X49" ~
          "Accidental poisoning",

        icd10_3 >= "X60" & icd10_3 <= "X84" ~
          "Suicide",

        (icd10_3 >= "X85" & icd10_3 <= "X99") |
          (icd10_3 >= "Y00" & icd10_3 <= "Y09") ~
          "Homicide",

        icd10_3 >= "Y10" & icd10_3 <= "Y34" ~
          "Undetermined intent",

        str_detect(underlying_cause, "^[VWXY]") ~
          "Other external causes",

        TRUE ~
          NA_character_
      )
  )

# Minimal quality checks required for the analysis.
stopifnot(
  all(women_deaths$death_year[!is.na(women_deaths$death_year)] %in% years),
  sum(!is.na(women_deaths$underlying_cause)) > 0
)
```

## Female population denominator

Annual female population denominators come from the **IBGE Population Projections, 2024 Revision**. Population refers to July 1 of each year. The same denominator is used for all national female mortality rates in this report.

```{r ibge-female-population, results='hide'}
ibge_population_url <-
  paste0(
    "https://ftp.ibge.gov.br/Projecao_da_Populacao/",
    "Projecao_da_Populacao_2024/",
    "projecoes_2024_tab1_idade_simples.xlsx"
  )

ibge_population_file <-
  file.path(
    population_directory,
    "projecoes_2024_tab1_idade_simples.xlsx"
  )

if (!file.exists(ibge_population_file)) {
  download.file(
    url = ibge_population_url,
    destfile = ibge_population_file,
    mode = "wb",
    quiet = FALSE
  )
}

ibge_population_raw <-
  read_excel(
    ibge_population_file,
    sheet = "1) POP_IDADE SIMPLES",
    skip = 5
  )

population_year_columns <-
  as.character(2010:2024)

stopifnot(
  all(population_year_columns %in% names(ibge_population_raw))
)

female_population_by_year <-
  ibge_population_raw %>%
  filter(
    SIGLA == "BR",
    SEXO == "Mulheres"
  ) %>%
  select(
    IDADE,
    all_of(population_year_columns)
  ) %>%
  pivot_longer(
    cols = all_of(population_year_columns),
    names_to = "death_year",
    values_to = "female_population"
  ) %>%
  mutate(
    death_year = as.integer(death_year),
    female_population = as.numeric(female_population)
  ) %>%
  group_by(death_year) %>%
  summarise(
    female_population =
      sum(
        female_population,
        na.rm = TRUE
      ),
    .groups = "drop"
  ) %>%
  arrange(death_year)

stopifnot(
  nrow(female_population_by_year) == 15,
  identical(female_population_by_year$death_year, 2010:2024),
  all(female_population_by_year$female_population > 0)
)
```

# Main results

## 1. Female deaths: absolute numbers

The analysis begins with the annual number of deaths among women recorded in SIM, irrespective of cause.

```{r female-deaths-absolute, fig.width=10, fig.height=6.2}
deaths_by_year <-
  women_deaths %>%
  count(
    death_year,
    name = "Deaths"
  ) %>%
  filter(death_year %in% years) %>%
  arrange(death_year)

plot_deaths_over_time <-
  ggplot(
    deaths_by_year,
    aes(
      x = death_year,
      y = Deaths
    )
  ) +
  geom_line(
    linewidth = 1.1,
    color = col_blue
  ) +
  geom_point(
    size = 2.6,
    color = col_blue
  ) +
  scale_x_continuous(
    breaks = years
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.03, 0.08))
  ) +
  labs(
    title = "Deaths among women recorded in SIM",
    subtitle = "Annual number of deaths, Brazil, 2010–2024",
    x = "Year",
    y = "Number of deaths",
    caption =
      "Source: Brazilian Mortality Information System (SIM/DATASUS)."
  ) +
  theme_datasus() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

plot_deaths_over_time <- add_contact_footer(plot_deaths_over_time)

plot_deaths_over_time
```

Absolute counts describe the number of recorded deaths, but they do not account for changes in the size of the female population.

## 2. Female deaths: crude mortality rate

To account for population size, annual deaths are divided by the female population of Brazil. The resulting measure is a **crude mortality rate per 100,000 women**; it is not age-standardized.

```{r female-crude-mortality-rate, fig.width=10, fig.height=6.2}
female_mortality_rate <-
  deaths_by_year %>%
  transmute(
    death_year = as.integer(death_year),
    female_deaths = as.numeric(Deaths)
  ) %>%
  left_join(
    female_population_by_year,
    by = "death_year"
  ) %>%
  mutate(
    deaths_per_100000_women =
      100000 * female_deaths / female_population
  )

stopifnot(
  !any(is.na(female_mortality_rate$deaths_per_100000_women))
)

plot_female_crude_mortality <-
  ggplot(
    female_mortality_rate,
    aes(
      x = death_year,
      y = deaths_per_100000_women
    )
  ) +
  geom_line(
    linewidth = 1.15,
    color = col_blue
  ) +
  geom_point(
    size = 2.7,
    color = col_blue
  ) +
  scale_x_continuous(
    breaks = years
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 1),
    expand = expansion(mult = c(0.04, 0.08))
  ) +
  labs(
    title = "Crude mortality rate among women in Brazil",
    subtitle = "Deaths from all causes per 100,000 women, 2010–2024",
    x = "Year",
    y = "Deaths per 100,000 women",
    caption = paste0(
      "Deaths: Mortality Information System (SIM/DATASUS). ",
      "Population: IBGE Population Projections, 2024 Revision.\n",
      "Population corresponds to women in Brazil on July 1 of each year. ",
      "Crude rates are not adjusted for age."
    )
  ) +
  theme_datasus() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

plot_female_crude_mortality <- add_contact_footer(plot_female_crude_mortality)

plot_female_crude_mortality
```

## 3. External and violent causes of death

The focus now narrows from mortality from all causes to deaths classified within ICD-10 external causes. Homicide, suicide and events of undetermined intent are kept as distinct categories.

```{r external-causes, fig.width=10, fig.height=6.2}
external_cause_summary <-
  women_deaths %>%
  filter(
    !is.na(external_cause_category)
  ) %>%
  count(
    external_cause_category,
    name = "Deaths"
  ) %>%
  arrange(desc(Deaths))

plot_external_causes <-
  external_cause_summary %>%
  mutate(
    external_cause_category =
      fct_reorder(
        external_cause_category,
        Deaths
      )
  ) %>%
  ggplot(
    aes(
      x = external_cause_category,
      y = Deaths
    )
  ) +
  geom_col(
    fill = col_gold,
    width = 0.72
  ) +
  coord_flip() +
  geom_text(
    aes(label = scales::comma(Deaths)),
    hjust = -0.12,
    size = 3.5,
    color = "#334E5C"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "External causes of death among women",
    subtitle = "Absolute numbers, Brazil, 2010–2024",
    x = NULL,
    y = "Number of deaths",
    caption =
      "Source: Brazilian Mortality Information System (SIM/DATASUS)."
  ) +
  theme_datasus() +
  theme(
    panel.grid.major.y = element_blank()
  )

plot_external_causes <- add_contact_footer(plot_external_causes)

plot_external_causes
```

<div class="warning-box">
<strong>Important distinction.</strong> Suicide is defined as ICD-10 X60–X84 and homicide/assault as X85–Y09. Events of undetermined intent (Y10–Y34) remain separate. SIM identifies deaths of women classified as homicide, but the underlying ICD-10 cause alone does not contain the legal and contextual information required to establish femicide.
</div>

## 4. Homicide and suicide: absolute numbers

The next step isolates the two intentional violent causes examined in the main analysis: homicide and suicide.

```{r homicide-suicide-data, results='hide'}
suicide_homicide_year <-
  women_deaths %>%
  filter(
    external_cause_category %in%
      c("Suicide", "Homicide")
  ) %>%
  count(
    death_year,
    external_cause_category,
    name = "deaths"
  ) %>%
  tidyr::complete(
    death_year = years,
    external_cause_category = c("Suicide", "Homicide"),
    fill = list(deaths = 0)
  ) %>%
  arrange(death_year, external_cause_category)

suicide_homicide_total <-
  suicide_homicide_year %>%
  group_by(external_cause_category) %>%
  summarise(
    deaths = sum(deaths),
    .groups = "drop"
  ) %>%
  mutate(
    percentage = 100 * deaths / sum(deaths)
  )

total_suicide <-
  suicide_homicide_total %>%
  filter(external_cause_category == "Suicide")

total_homicide <-
  suicide_homicide_total %>%
  filter(external_cause_category == "Homicide")

total_violent <-
  sum(suicide_homicide_total$deaths)
```

```{r homicide-suicide-absolute, fig.width=10, fig.height=6.4}
plot_homicide_suicide_absolute <-
  ggplot(
    suicide_homicide_year,
    aes(
      x = death_year,
      y = deaths,
      color = external_cause_category,
      group = external_cause_category
    )
  ) +
  geom_line(linewidth = 1.15) +
  geom_point(size = 2.6) +
  scale_color_manual(
    values = c(
      "Homicide" = col_red,
      "Suicide" = col_teal
    )
  ) +
  scale_x_continuous(
    breaks = years
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.03, 0.08))
  ) +
  labs(
    title = "Homicide and suicide deaths among women in Brazil",
    subtitle = "Annual absolute number of deaths, 2010–2024",
    x = "Year",
    y = "Number of deaths",
    color = NULL,
    caption = paste0(
      "2010–2024 total: ", scales::comma(total_violent),
      " homicide + suicide deaths. Homicide: ",
      scales::comma(total_homicide$deaths),
      "; Suicide: ", scales::comma(total_suicide$deaths), ".\n",
      "Source: Brazilian Mortality Information System (SIM/DATASUS)."
    )
  ) +
  theme_datasus() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

plot_homicide_suicide_absolute <- add_contact_footer(plot_homicide_suicide_absolute)

plot_homicide_suicide_absolute
```

## 5. Homicide and suicide: mortality rates

Absolute numbers are then converted to crude cause-specific mortality rates using the same annual female population denominator from IBGE.

```{r homicide-suicide-rates, fig.width=10, fig.height=6.4}
female_homicide_suicide_rates <-
  suicide_homicide_year %>%
  transmute(
    death_year = as.integer(death_year),
    cause = as.character(external_cause_category),
    deaths = as.numeric(deaths)
  ) %>%
  left_join(
    female_population_by_year,
    by = "death_year"
  ) %>%
  mutate(
    deaths_per_100000_women =
      100000 * deaths / female_population
  ) %>%
  arrange(death_year, cause)

stopifnot(
  nrow(female_homicide_suicide_rates) == 30,
  !any(is.na(female_homicide_suicide_rates$deaths_per_100000_women))
)

plot_female_homicide_suicide_rates <-
  ggplot(
    female_homicide_suicide_rates,
    aes(
      x = death_year,
      y = deaths_per_100000_women,
      color = cause,
      group = cause
    )
  ) +
  geom_line(linewidth = 1.15) +
  geom_point(size = 2.7) +
  scale_color_manual(
    values = c(
      "Homicide" = col_red,
      "Suicide" = col_teal
    )
  ) +
  scale_x_continuous(
    breaks = years
  ) +
  scale_y_continuous(
    labels = scales::label_number(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.10))
  ) +
  labs(
    title = "Homicide and suicide mortality rates among women in Brazil",
    subtitle = "Deaths per 100,000 women, 2010–2024",
    x = "Year",
    y = "Deaths per 100,000 women",
    color = NULL,
    caption = paste0(
      "Deaths: Mortality Information System (SIM/DATASUS). ",
      "Population: IBGE Population Projections, 2024 Revision.\n",
      "Crude cause-specific mortality rates are not adjusted for age."
    )
  ) +
  theme_datasus() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

plot_female_homicide_suicide_rates <- add_contact_footer(plot_female_homicide_suicide_rates)

plot_female_homicide_suicide_rates
```

## 6. Femicide: registered cases and relation to female homicides

Brazil introduced the legal category of femicide in 2015. The series below uses the consolidated annual counts published by the **Fórum Brasileiro de Segurança Pública (FBSP)**. These records originate from public-security data and are therefore analytically distinct from deaths classified as homicide in SIM.

The first figure compares the absolute number of registered femicides with female homicide deaths recorded in SIM. The second expresses the FBSP femicide count relative to the SIM female-homicide count in the same year.

```{r femicide-data, results='hide'}
# Consolidated FBSP historical series used in this project.
femicide_fbsp <-
  tibble::tibble(
    death_year = 2015:2024,
    femicides_fbsp = c(
      449,
      929,
      1075,
      1229,
      1330,
      1354,
      1347,
      1455,
      1475,
      1492
    )
  )

female_homicide_sim <-
  suicide_homicide_year %>%
  filter(
    external_cause_category == "Homicide",
    death_year >= 2015,
    death_year <= 2024
  ) %>%
  transmute(
    death_year = as.integer(death_year),
    female_homicides_sim = as.numeric(deaths)
  )

femicide_homicide_comparison <-
  female_homicide_sim %>%
  left_join(
    femicide_fbsp,
    by = "death_year"
  ) %>%
  mutate(
    femicide_percentage_of_sim_homicides =
      100 * femicides_fbsp / female_homicides_sim
  ) %>%
  arrange(death_year)
```

### Registered femicides and female homicides

```{r femicide-absolute, fig.width=10, fig.height=6.4}
plot_femicide_absolute_data <-
  femicide_homicide_comparison %>%
  select(
    death_year,
    female_homicides_sim,
    femicides_fbsp
  ) %>%
  pivot_longer(
    cols = c(
      female_homicides_sim,
      femicides_fbsp
    ),
    names_to = "series",
    values_to = "deaths"
  ) %>%
  mutate(
    series =
      recode(
        series,
        female_homicides_sim = "Female homicides — SIM/DATASUS",
        femicides_fbsp = "Registered femicides — FBSP"
      )
  )

plot_femicide_absolute <-
  ggplot(
    plot_femicide_absolute_data,
    aes(
      x = death_year,
      y = deaths,
      color = series,
      group = series
    )
  ) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.6) +
  scale_color_manual(
    values = c(
      "Female homicides — SIM/DATASUS" = col_navy,
      "Registered femicides — FBSP" = col_red
    )
  ) +
  scale_x_continuous(
    breaks = 2015:2024
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.03, 0.08))
  ) +
  labs(
    title = "Female homicides and registered femicides in Brazil",
    subtitle = "SIM/DATASUS homicide deaths and FBSP femicide records, 2015–2024",
    x = "Year",
    y = "Number of deaths",
    color = NULL,
    caption = paste0(
      "Female homicides: Mortality Information System (SIM/DATASUS). ",
      "Femicides: Fórum Brasileiro de Segurança Pública (FBSP).\n",
      "The two series originate from different information systems."
    )
  ) +
  theme_datasus()

plot_femicide_absolute <- add_contact_footer(plot_femicide_absolute)

plot_femicide_absolute
```

### Registered femicides relative to female homicides

```{r femicide-proportion, fig.width=10, fig.height=6.4}
plot_femicide_percentage <-
  ggplot(
    femicide_homicide_comparison,
    aes(
      x = death_year,
      y = femicide_percentage_of_sim_homicides
    )
  ) +
  geom_line(
    color = col_red,
    linewidth = 1.1
  ) +
  geom_point(
    color = col_red,
    size = 2.6
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.1f%%",
        femicide_percentage_of_sim_homicides
      )
    ),
    vjust = -0.8,
    size = 3.2,
    color = "#343A40"
  ) +
  scale_x_continuous(
    breaks = 2015:2024
  ) +
  scale_y_continuous(
    labels = scales::label_percent(
      scale = 1,
      accuracy = 1
    ),
    expand = expansion(mult = c(0.03, 0.15))
  ) +
  labs(
    title = "Registered femicides relative to female homicides in Brazil",
    subtitle = "FBSP femicides as a percentage of female homicide deaths recorded in SIM, 2015–2024",
    x = "Year",
    y = "Femicides / female homicides (%)",
    caption = paste0(
      "Numerator: registered femicides, FBSP. ",
      "Denominator: female homicide deaths, SIM/DATASUS.\n",
      "This is a cross-system descriptive comparison, not the official FBSP proportion."
    )
  ) +
  theme_datasus()

plot_femicide_percentage <- add_contact_footer(plot_femicide_percentage)

plot_femicide_percentage
```

The ratio above should be interpreted literally as **registered femicides in FBSP divided by female homicide deaths recorded in SIM**. Because numerator and denominator come from different information systems, it should not be interpreted as an individual-level classification of SIM homicide records.

# Additional demographic analyses

These figures are retained for follow-up analyses, including a separate communication focused on differences by race/skin color. They show **absolute counts**, not subgroup-specific mortality rates.

## Female deaths by age group

```{r age-plot, fig.width=10, fig.height=5.8}
deaths_by_age_group <-
  women_deaths %>%
  filter(!is.na(age_group)) %>%
  count(
    age_group,
    name = "Deaths"
  ) %>%
  mutate(
    Percentage = 100 * Deaths / sum(Deaths)
  )

plot_age_groups <-
  ggplot(
    deaths_by_age_group,
    aes(
      x = age_group,
      y = Deaths
    )
  ) +
  geom_col(
    fill = col_blue,
    width = 0.72
  ) +
  geom_text(
    aes(
      label = paste0(
        scales::comma(Deaths),
        "\n(", sprintf("%.1f", Percentage), "%)"
      )
    ),
    vjust = -0.25,
    size = 3.1,
    color = "#334E5C"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Age distribution of deaths among women",
    subtitle = "Absolute numbers, Brazil, 2010–2024",
    x = "Age group",
    y = "Number of deaths",
    caption =
      "Source: Brazilian Mortality Information System (SIM/DATASUS)."
  ) +
  theme_datasus() +
  theme(
    panel.grid.major.x = element_blank()
  )

plot_age_groups <- add_contact_footer(plot_age_groups)

plot_age_groups
```

## Female deaths by race/skin color

```{r race-plot, fig.width=10, fig.height=5.8}
deaths_by_race <-
  women_deaths %>%
  mutate(
    race_skin_color =
      if_else(
        is.na(RACACOR) |
          str_trim(as.character(RACACOR)) == "",
        "Missing/unknown",
        as.character(RACACOR)
      )
  ) %>%
  count(
    race_skin_color,
    name = "Deaths"
  ) %>%
  arrange(desc(Deaths))

plot_deaths_by_race <-
  deaths_by_race %>%
  mutate(
    race_skin_color =
      fct_reorder(
        race_skin_color,
        Deaths
      )
  ) %>%
  ggplot(
    aes(
      x = race_skin_color,
      y = Deaths
    )
  ) +
  geom_col(
    fill = col_blue,
    width = 0.72
  ) +
  coord_flip() +
  geom_text(
    aes(label = scales::comma(Deaths)),
    hjust = -0.10,
    size = 3.3,
    color = "#334E5C"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.13))
  ) +
  labs(
    title = "Deaths among women by race/skin color",
    subtitle = "Absolute numbers, Brazil, 2010–2024",
    x = NULL,
    y = "Number of deaths",
    caption = paste0(
      "Counts are not race-specific mortality rates. ",
      "Missing/unknown values remain explicit.\n",
      "Source: Brazilian Mortality Information System (SIM/DATASUS)."
    )
  ) +
  theme_datasus() +
  theme(
    panel.grid.major.y = element_blank()
  )

plot_deaths_by_race <- add_contact_footer(plot_deaths_by_race)

plot_deaths_by_race
```

## Homicide and suicide by age group

```{r homicide-suicide-by-age, fig.width=10, fig.height=6.2}
suicide_homicide_age <-
  women_deaths %>%
  filter(
    external_cause_category %in% c("Suicide", "Homicide"),
    !is.na(age_group)
  ) %>%
  count(
    age_group,
    external_cause_category,
    name = "deaths"
  ) %>%
  tidyr::complete(
    age_group,
    external_cause_category = c("Suicide", "Homicide"),
    fill = list(deaths = 0)
  )

plot_homicide_suicide_by_age <-
  ggplot(
    suicide_homicide_age,
    aes(
      x = age_group,
      y = deaths,
      fill = external_cause_category
    )
  ) +
  geom_col(
    position = "dodge",
    width = 0.72
  ) +
  scale_fill_manual(
    values = c(
      "Homicide" = col_red,
      "Suicide" = col_teal
    )
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = "Homicide and suicide deaths among women by age group",
    subtitle = "Absolute numbers, Brazil, 2010–2024",
    x = "Age group",
    y = "Number of deaths",
    fill = NULL,
    caption = paste0(
      "Counts are not age-specific mortality rates.\n",
      "Source: Brazilian Mortality Information System (SIM/DATASUS)."
    )
  ) +
  theme_datasus() +
  theme(
    panel.grid.major.x = element_blank()
  )

plot_homicide_suicide_by_age <- add_contact_footer(plot_homicide_suicide_by_age)

plot_homicide_suicide_by_age
```

## Homicide and suicide by race/skin color

```{r homicide-suicide-by-race, fig.width=10, fig.height=6.2}
suicide_homicide_race <-
  women_deaths %>%
  filter(
    external_cause_category %in% c("Suicide", "Homicide")
  ) %>%
  mutate(
    race_skin_color =
      if_else(
        is.na(RACACOR) |
          str_trim(as.character(RACACOR)) == "",
        "Missing/unknown",
        as.character(RACACOR)
      )
  ) %>%
  count(
    race_skin_color,
    external_cause_category,
    name = "deaths"
  )

plot_homicide_suicide_by_race <-
  ggplot(
    suicide_homicide_race,
    aes(
      x = fct_reorder(
        race_skin_color,
        deaths,
        .fun = sum
      ),
      y = deaths,
      fill = external_cause_category
    )
  ) +
  geom_col(
    position = "dodge",
    width = 0.72
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Homicide" = col_red,
      "Suicide" = col_teal
    )
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.10))
  ) +
  labs(
    title = "Homicide and suicide deaths among women by race/skin color",
    subtitle = "Absolute numbers, Brazil, 2010–2024",
    x = NULL,
    y = "Number of deaths",
    fill = NULL,
    caption = paste0(
      "Counts are not race-specific mortality rates. ",
      "Missing/unknown values remain explicit.\n",
      "Source: Brazilian Mortality Information System (SIM/DATASUS)."
    )
  ) +
  theme_datasus() +
  theme(
    panel.grid.major.y = element_blank()
  )

plot_homicide_suicide_by_race <- add_contact_footer(plot_homicide_suicide_by_race)

plot_homicide_suicide_by_race
```

# Interpretation and limitations

The report is descriptive. National crude mortality rates account for annual changes in the size of the female population, but they are **not age-standardized**. The age and race/skin-color figures are distributions of recorded deaths and should not be interpreted as subgroup mortality risks without corresponding subgroup population denominators.

For external causes, classification depends on the underlying cause recorded in SIM. Deaths of undetermined intent (ICD-10 Y10–Y34) are kept separate rather than reassigned to homicide or suicide. This is important because changes in cause-of-death classification can affect observed trends.

The femicide analysis combines two distinct official/administrative information systems. SIM provides female homicide deaths based on death certificates and ICD-10 underlying cause, whereas FBSP compiles femicide records from public-security sources. Consequently, the calculated **FBSP femicides / SIM female homicides** percentage is a cross-system descriptive comparison and is not the official FBSP proportion of female homicides classified as femicide.

# Reproducibility and exports

Only the analytical datasets and figures used in the streamlined report are exported below.

```{r export-results, results='hide'}
# Analytical tables.
write_csv(
  deaths_by_year,
  file.path(output_directory, "female_deaths_by_year_2010_2024.csv")
)

write_csv(
  female_mortality_rate,
  file.path(output_directory, "female_crude_mortality_rate_per_100000_2010_2024.csv")
)

write_csv(
  external_cause_summary,
  file.path(output_directory, "external_causes_women_2010_2024.csv")
)

write_csv(
  suicide_homicide_year,
  file.path(output_directory, "female_homicide_suicide_by_year_2010_2024.csv")
)

write_csv(
  female_homicide_suicide_rates,
  file.path(output_directory, "female_homicide_suicide_rates_per_100000_2010_2024.csv")
)

write_csv(
  femicide_homicide_comparison,
  file.path(output_directory, "femicide_vs_female_homicide_2015_2024.csv")
)

write_csv(
  suicide_homicide_age,
  file.path(output_directory, "female_homicide_suicide_by_age_2010_2024.csv")
)

write_csv(
  suicide_homicide_race,
  file.path(output_directory, "female_homicide_suicide_by_race_2010_2024.csv")
)

# Main figures.
ggsave(
  file.path(figure_directory, "01_female_deaths_absolute.png"),
  plot_deaths_over_time,
  width = 10,
  height = 6.2,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "02_female_mortality_rate_per_100000.png"),
  plot_female_crude_mortality,
  width = 10,
  height = 6.2,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "03_external_causes_women.png"),
  plot_external_causes,
  width = 10,
  height = 6.2,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "04_homicide_suicide_absolute.png"),
  plot_homicide_suicide_absolute,
  width = 10,
  height = 6.4,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "05_homicide_suicide_rates_per_100000.png"),
  plot_female_homicide_suicide_rates,
  width = 10,
  height = 6.4,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "06_femicide_and_female_homicide_absolute.png"),
  plot_femicide_absolute,
  width = 10,
  height = 6.4,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "07_femicide_relative_to_sim_homicides.png"),
  plot_femicide_percentage,
  width = 10,
  height = 6.4,
  dpi = 320,
  bg = "white"
)

# Additional demographic figures.
ggsave(
  file.path(figure_directory, "08_female_deaths_by_age.png"),
  plot_age_groups,
  width = 10,
  height = 5.8,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "09_female_deaths_by_race.png"),
  plot_deaths_by_race,
  width = 10,
  height = 5.8,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "10_homicide_suicide_by_age.png"),
  plot_homicide_suicide_by_age,
  width = 10,
  height = 6.2,
  dpi = 320,
  bg = "white"
)

ggsave(
  file.path(figure_directory, "11_homicide_suicide_by_race.png"),
  plot_homicide_suicide_by_race,
  width = 10,
  height = 6.2,
  dpi = 320,
  bg = "white"
)

# Reproducibility information.
writeLines(
  capture.output(sessionInfo()),
  file.path(output_directory, "sessionInfo.txt")
)
```
